---
title: "Advent of Code 2020"
output: html_notebook
---

```{r}
library(tidyverse)
```

#--- Day 1: Report Repair ---

Find the two entries that sum to 2020; what do you get if you multiply them together?

*Strategy*: basically brute force, sum every combination

```{r}
day1_input <- scan("day1_input.txt") 

expand_grid(x = day1_input, y = day1_input) %>%
  rowwise() %>% 
  mutate(sum_xy = sum(x, y)) %>% 
  filter(sum_xy == 2020)

1683 * 337
```

#--- Part Two ---

In your expense report, what is the product of the three entries that sum to 2020?

*Strategy*: same thing, not great but it works.

```{r}
expand_grid(x = day1_input, y = day1_input, z = day1_input) %>%
  rowwise() %>% 
  mutate(sum_xyz = sum(x, y, z)) %>% 
  filter(sum_xyz == 2020)

862 * 877 * 281
```

#--- Day 2: Password Philosophy ---

How many passwords are valid according to their policies?

*Strategy*: Split each string into rule, letter to search for, and password. Then count `letter` in `password` and filter by rule

```{r}
day2_input <- read_csv("day2_input.txt", col_names = FALSE)

day2_data <- day2_input %>% 
  separate(col = X1, sep = " ", into = c("range", "letter", "password")) %>%
  separate(col = range, sep = "-", into = c("min_rule", "max_rule"), convert = TRUE) %>% 
  mutate(letter = str_replace(letter, "\\:$", ""))

day2_data %>% 
  mutate(letter_count = str_count(password, letter)) %>% 
  filter(letter_count >= min_rule & letter_count <= max_rule)
```

#--- Part Two ---

How many passwords are valid according to the new interpretation of the policies?

*Strategy*: pull out the specified letters from each password, collapse into single string and search for 1 occurrence of `letter`

```{r}
day2_input %>% 
  separate(col = X1, sep = " ", into = c("range", "letter", "password")) %>%
  separate(col = range, sep = "-", into = c("pos1", "pos2"), convert = TRUE) %>% 
  mutate(letter = str_replace(letter, "\\:$", "")) %>% 
  mutate(pass1 = str_sub(password, pos1, pos1),
         pass2 = str_sub(password, pos2, pos2)) %>% 
  unite("search_string", pass1:pass2, sep = "") %>% 
  filter(str_count(search_string, letter) == 1)
```

#--- Day 3: Toboggan Trajectory ---

Starting at the top-left corner of your map and following a slope of right 3 and down 1, how many trees would you encounter?

*Strategy*: separate map into rows and columns of a dataframe. Starting from [1,1], read the next cell (specified by adding [1,3] to position)  and add to a tree counter if `#`. Handle the case where if you go off the right side of the map (past col 31), you start over on the left.

```{r}
day3_data <- read_csv("day3_input.txt", col_names = FALSE) %>% 
  separate(col = X1, into = paste0("col", 1:32), sep = c(1:32)) %>% 
  select(-col32)
```

```{r}
rowpos = 1
colpos = 1
treecount = 0

while (rowpos <= 322) {
  
  rowpos = rowpos + 1
  
  if ((colpos + 3) > 31) {
    colpos = 0 + ((colpos+3) - 31)
  }else{
    colpos = colpos + 3
  }
  
  if(day3_data[rowpos, colpos] == "#") treecount = treecount + 1
  
}

treecount
```

#--- Part Two ---

What do you get if you multiply together the number of trees encountered on each of the listed slopes?

*Strategy*: same as above, but make it a function that takes slope as input.

```{r}
treecalc <- function(right, down) { 
  rowpos = 1
  colpos = 1
  treecount = 0
  
  while (rowpos <= 322) {
    
    rowpos = rowpos + down
    
    if ((colpos + right) > 31) {
      colpos = 0 + ((colpos + right) - 31)
    }else{
      colpos = colpos + right
    }
    
    if(day3_data[rowpos, colpos] == "#") treecount = treecount + 1
    
  }
  
  treecount
  
}
```

```{r}
treecalc(3,1) * treecalc(1,1) * treecalc(5,1) * treecalc(7,1) * treecalc(1,2)
```

#--- Day 4: Passport Processing ---

Count the number of valid passports - those that have all required fields. Treat cid as optional. In your batch file, how many passports are valid?

*Strategy*: use `read_file` to read in text file as a single string, then split into rows on "\n\n". Add a `passport_id` to group by, then separate out fields and values. Delete `cid` rows.  

```{r}
read_csv("day4_input.txt", col_names = FALSE, skip_empty_rows = FALSE) %>% 
  filter(is.na(X1)) %>% 
  nrow()

# should be 251 passports
```

```{r}
day4_data <- read_file("day4_input.txt") %>% 
  str_split('\\n{2,}') %>% 
  unlist() %>% 
  as_tibble_col(column_name = "passport") %>% 
  mutate(passport_id = row_number(), .before = 1) %>% 
  separate(col = passport, 
           into = paste0("field", 1:8),
           sep = "[[:space:]]",
           remove = FALSE)
```

```{r}
day4_data_long <- day4_data %>% 
  pivot_longer(cols = starts_with("field"),
               names_to = "field",
               values_to = "value") %>% 
  select(-field, -passport) %>% 
  separate(value,
           into = c("field", "value"),
           sep = ":") %>% 
  group_by(passport_id) %>%
  drop_na() %>%
  filter(field != "cid") %>% 
  ungroup()

day4_data_long %>%
  group_by(passport_id) %>% 
  tally() %>% 
  filter(n == 7)
```

# --- Part Two ---
Count the number of valid passports - those that have all required fields and valid values. Continue to treat cid as optional. In your batch file, how many passports are valid?

*Strategy*: lots of `case_when`. Helpful to separately filter for each field to get a sense of what the invalid data looks like.

(slight cheat: looked at this person's code and realized one of my regex's was wrong (but close): https://github.com/clanfear/advent_2020/commit/982230b9942f1197f3bcfb1e4a77228a56e329d0)

```{r}
day4_data_long %>%
  mutate(valid = case_when(field == "byr" & between(as.numeric(value), 1920, 2002) ~ "y",
                           field == "iyr" & between(as.numeric(value), 2010, 2020) ~ "y",
                           field == "eyr" & between(as.numeric(value), 2020, 2030) ~ "y",
                           field == "hgt" & str_detect(value, "cm") & between(as.numeric(str_sub(value, 1, -3)), 150, 193) ~ "y",
                           field == "hgt" & str_detect(value, "in") & between(as.numeric(str_sub(value, 1, -3)), 59, 76) ~ "y",
                           field == "hcl" & str_detect(value, "^#[0-9a-f]{6}$") ~ "y",
                           field == "ecl" & value %in% c("amb", "blu", "brn", "gry", "grn", "hzl", "oth") ~ "y",
                           field == "pid" & str_detect(value, "^[0-9]{9}$") ~ "y",
                           TRUE ~ "n")
         ) %>%
  filter(valid == "y") %>% 
  group_by(passport_id) %>% 
  tally() %>% 
  filter(n == 7)
```

#--- Day 5: Binary Boarding ---
As a sanity check, look through your list of boarding passes. What is the highest seat ID on a boarding pass?


#*Pre-work*: convert boarding pass string to vector, then iterate through each element
#
# ```{r}
# rows <- c(0:127)
# cols <- c(0:7)
# 
# #F (0-63)
# plane_rows[1:(length(plane_rows)/2)]
# 
# #B (64-127)
# plane_rows[-1:-(length(plane_rows)/2)]
# 
# #L (0-3)
# plane_cols[1:(length(plane_cols)/2)]
# 
# #R (4-7)
# plane_cols[-1:-(length(plane_cols)/2)]
# ```
# 
# ```{r}
# find_row <- function(input){
#   if (input == "F") {
#     rows <- rows[1:(length(rows)/2)]
#   }else if (input == "B"){
#     rows <- rows[-1:-(length(rows)/2)]
#   }
# }
# ```
# 
# Figure things out manually
# 
# ```{r}
# test <- "FBFBBFF" %>% 
#   str_split_fixed(., "", n = nchar(.)) %>% 
#   as.vector()
# 
# rows <- c(0:127)
# 
# rows <- find_row("F")
# rows <- find_row("B")
# rows <- find_row("F")
# rows <- find_row("B")
# rows <- find_row("B")
# rows <- find_row("F")
# rows <- find_row("F")
# rows
# ```
# ```{r}
# rows <- c(0:127)
# i <- 1
# rows <- find_row(test[i])
# rows <- find_row(test[i+1])
# rows <- find_row(test[i+2])
# rows <- find_row(test[i+3])
# rows <- find_row(test[i+4])
# rows <- find_row(test[i+5])
# rows <- find_row(test[i+6])
# rows
# ```
# 
# ```{r}
# rows <- c(0:127)
# 
# for (i in 1:length(test)) {
#   rows <- find_row(test[i])
#   i <- i+1
#   }
# 
# rows
# ```
# 
# ```{r}
# find_col <- function(input){
#   if (input == "L") {
#     cols <- cols[1:(length(cols)/2)]
#   }else if (input == "R"){
#     cols <- cols[-1:-(length(cols)/2)]
#   }
# }
# ```
# ```{r}
# testcols <- "RLR" %>% 
#   str_split_fixed(., "", n = nchar(.)) %>% 
#   as.vector()
# 
# cols <- c(0:7)
# i <- 1
# cols <- find_col(testcols[i])
# cols <- find_col(testcols[i+1])
# cols <- find_col(testcols[i+2])
# cols
# ```
# 
# ```{r}
# cols <- c(0:7)
# 
# for (i in 1:length(testcols)) {
#   cols <- find_col(testcols[i])
#   i <- i+1
#   }
# 
# cols
# ```

**Start of actual solution**

Read in boarding passes, split into row and col instructions

```{r}
day5_data <- read_csv("day5_input.txt", col_names = "boarding_pass") %>% 
  separate(col = boarding_pass,
           into = c("row_instructions", "col_instructions"),
           sep = -3,
           remove = FALSE)
```

Function that identifies column from instructions  

```{r}
cols_loop <- function(instructions){
  cols <- c(0:7)
  
  find_col <- function(input){
    if (input == "L") {
      cols <- cols[1:(length(cols)/2)]
    }else if (input == "R"){
      cols <- cols[-1:-(length(cols)/2)]
    }
  }
  
  instructions <- str_split_fixed(instructions, "", 3)
  
  for (i in 1:length(instructions)) {
    cols <- find_col(instructions[i])
    i <- i+1
  }
  return(cols)
}
```

Function that identifies row from instructions

```{r}
rows_loop <- function(instructions){
  rows <- c(0:127)
  
  find_row <- function(input){
    if (input == "F") {
      rows <- rows[1:(length(rows)/2)]
    }else if (input == "B"){
      rows <- rows[-1:-(length(rows)/2)]
    }
  }
  
  instructions <- str_split_fixed(instructions, "", 7)
  
  for (i in 1:length(instructions)) {
    rows <- find_row(instructions[i])
    i <- i+1
  }
  return(rows)
}
```

Find row and column from each set of instructions, then calculate seat id for each row.

```{r}
day5_output <- day5_data %>% 
  mutate(col_output = map_dbl(col_instructions, cols_loop),
         row_output = map_dbl(row_instructions, rows_loop)) %>% 
  rowwise() %>% 
  mutate(seat_id = row_output * 8 + col_output) %>% 
  arrange(desc(seat_id))
```

#--- Part Two ---
What is the ID of your seat?

*Strategy:* set up a dataframe with all row/col combos and compare to list of boarding passes 

```{r}
airplane <- expand_grid(row_output = c(0:127), col_output = c(0:7))

board_list <- day5_output %>% select(row_output, col_output)

# rows in airplane that are not in board_list
setdiff(airplane, board_list) %>% 
  mutate(seat_id = row_output * 8 + col_output) %>% 
  arrange(seat_id)

# there is one seat in the middle that is by itself (the others are blocks at the front & back)
```

#--- Day 6: Custom Customs ---
As your flight approaches the regional airport where you'll switch to a much larger plane, customs declaration forms are distributed to the passengers.

The form asks a series of 26 yes-or-no questions marked a through z. All you need to do is identify the questions for which anyone in your group answers "yes". Since your group is just you, this doesn't take very long.

However, the person sitting next to you seems to be experiencing a language barrier and asks if you can help. For each of the people in their group, you write down the questions for which they answer "yes", one per line. For example:

abcx
abcy
abcz
In this group, there are 6 questions to which anyone answered "yes": a, b, c, x, y, and z. (Duplicate answers to the same question don't count extra; each question counts at most once.)

Another group asks for your help, then another, and eventually you've collected answers from every group on the plane (your puzzle input). Each group's answers are separated by a blank line, and within each group, each person's answers are on a single line. For example:

abc

a
b
c

ab
ac

a
a
a
a

b
This list represents answers from five groups:

The first group contains one person who answered "yes" to 3 questions: a, b, and c.
The second group contains three people; combined, they answered "yes" to 3 questions: a, b, and c.
The third group contains two people; combined, they answered "yes" to 3 questions: a, b, and c.
The fourth group contains four people; combined, they answered "yes" to only 1 question, a.
The last group contains one person who answered "yes" to only 1 question, b.
In this example, the sum of these counts is 3 + 3 + 3 + 1 + 1 = 11.

For each group, count the number of questions to which anyone answered "yes". What is the sum of those counts?

*Strategy*:

Read file as single string, split into groups by "\n\n", number groups, remove "\n", count distinct letters

```{r}
day6_data <- read_file("day6_input.txt") %>% 
  str_split('\\n{2,}') %>% 
  unlist() %>% 
  as_tibble_col(column_name = "responses") %>% 
  mutate(group_id = row_number(), .before = 1) 
```

```{r}
count_answers <- function(input){
  str_replace_all(input, "\n", "") %>% 
  str_split("") %>%
  unlist() %>% 
  n_distinct()
}

```

```{r}
day6_data %>% 
  mutate(distinct_responses = map_dbl(responses, count_answers), .before = 2) %>% 
  pull(distinct_responses) %>% 
  sum()
```

#--- Part Two ---

For each group, count the number of questions to which everyone answered "yes". What is the sum of those counts?

*Strategy:* leave each group as 1 string, count how many letters occur `n_people` times in each string.

Test once:

```{r}
test <- ("lznroxbqymvfijpwkec\ngljkpwyvsbmroziefnqxc\nbmkiewyxjfzqrocnlpv\nibewnmlkzfcrjyvxopq\nfkmxpbvjiwlrzocqyne")

n_people <- 1 + str_count(test, "\n")

letter_count <- 0

for (i in 1:length(letters)) {
  if(str_count(test, letters[i]) == n_people){
    letter_count <- letter_count + 1
  }
  i = i + 1
  
}

letter_count

```

Make function

```{r}
count_letters <- function(input) {
  
  n_people <- 1 + str_count(input, "\n")
  
  letter_count <- 0
  
  for (i in 1:length(letters)) {
    if(str_count(input, letters[i]) == n_people){
      letter_count <- letter_count + 1
    }
    i = i + 1
    
  }
  letter_count
}

```

Now apply to entire dataset

```{r}
day6_data %>% 
  #mutate(n_people = 1 + str_count(responses, "\n"), .after = 1) %>% 
  mutate(letter_counts = map_dbl(responses, count_letters), .before = 2) %>% 
  pull(letter_counts) %>% 
  sum()
```
